---
title: "Estimate weight cumulative exposure effect for binary outcome"
output: 
  rmarkdown::html_vignette:
    toc: true #table of content true
    toc_depth : 3
vignette: >
  %\VignetteIndexEntry{Estimate weight cumulative exposure effect for binary outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(mixWCE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

#pkgdown::build_site()
```

# background and definitions 

## Modeling the Exposure Trajectory

We assume the true continuous exposure $E_{i}^*(t)$ of subject $i$ at time $t$ follows a normal distribution with measurement noise. The underlying true exposure follows a linear mixed-effects model:

$$ E_{ij}=\underbrace{X_i(t)^\top\beta + Z_{i}(t)^\top b_i}_{E_i^*(t_{ij})}+\varepsilon_{ij} $$

where $X_i(t)$ is the vector of explanatory variables, $\beta$ the fixed effects,$Z_i(t)$ a subset of $X_i(t)$, and $b_i \sim N$(0,B)$ are subject-specific random effects. The time evolution is described either by a simple time function or by natural cubic splines to model more flexible variations.

## Event Modeling

We define the WCIE (weighted cumulative exposure) for subject $i$ as the weighted sum of the true exposure $E_{i}^*(t)$ over a time window $[-S,0]$ with unknown weights $w(t)$ and $S$ the time window exposure.

This cumulative exposure is then included in a logistic model to estimate the probability that the individual is a case ($Y_i=1$):

\begin{align*}
    \text{logit}(P(Y_i=1)) &= \alpha_0 + \alpha_pX_{ip} + {WCIE_i}\\
    &= \alpha_0 + \alpha_pX_{ip} + {\sum_{t=-S}^{0} E_i(t) \times w(t)}
    \label{eq7}
\end{align*}


where $\alpha_0$ is the intercept, $\alpha_p$ are the coefficients of covariates $X_{ip}$, and the sum represents the weighted cumulative effect of exposure.

## Modeling the Association

To estimate the exposure effects in the model, we first need to estimate the weight function $w(t)$. This function is expressed as a linear combination of basis functions:

$$w(t)= \sum_{k=0}^{K} \delta_k B_k(t)$$

where the coefficients $\delta_k$ are to be estimated and $B(k)$ a basis functions.

Estimating the exposure effects means estimating the coefficients $\delta_k$, which weight the contributions of each basis function in the logistic model.

This vignette illustrates how to estimate  weighted cumulative effect of a longitudinal exposition following the methodology described in Wagner et al. (2021 - https://doi.org/10.1186/s12874-021-01403-w). 

# Exemple for paquid data

In this example, we study the occurrence of physical dependency defined by the HIER2 score from the Paquid dataset. HIER is a physical dependency score with four levels: 0 (no dependency), 1 (mild dependency), 2 (moderate dependency), and 3 (severe dependency). For binary models, the outcome is defined as an indicator of moderate to severe dependency, i.e., a HIER score greater than or equal to 2. We investigate the effects of an exposure corresponding to the Mini-Mental State Examination (MMSE) score, a psychometric test of global cognitive functioning scored as an integer between 0 and 30. The sample includes elderly individuals followed over time, with adjustments possible for covariates. Interactions are not considered here.

## paquid subset

### exposition variable: normalized MMSE

As the MMSE score has a highly skewed distribution, we use its normalized version, which is easily obtained using the NormPsy package. This normalization is necessary to satisfy the assumptions of the linear mixed model used to describe the exposure trajectory, particularly the assumption that the residuals (and random effects) follow a normal distribution.

```{r}
library(NormPsy)
paquid$normMMSE <- normMMSE(paquid$MMSE)
```


### split dataset

In order to obtain our two time windows, we split the dataset in two sub dataset. The first one corrsponding to the outcome windows and keep one line per subject when he developp at least a moderate physical dependency (HIER >= 2), this event appear at $t=0$. The second dataset corresponding to the entire exposition window before this event $t=[-S,0]$.

```{r}
library(dplyr)
paquid_outcome <- paquid %>% 
  arrange(ID, time) %>%                # ordre chronologique
  group_by(ID) %>% 
  slice(
    if (any(HIER2 == 1, na.rm = TRUE)) {
      which(HIER2 == 1)[1]             # 1re visite avec HIER2 == 1
    } else {
      n()                               # sinon, dernière visite
    }
  ) %>% 
  ungroup()

paquid_expo <- paquid %>%
  arrange(ID, time) %>%
  group_by(ID) %>%
  mutate(
    time_outcome = if (any(HIER2 == 1, na.rm = TRUE)) {
      time[which(HIER2 == 1)[1]]  # 1er moment HIER2 == 1
    } else {
      max(time, na.rm = TRUE)     # sinon dernière visite
    }
  ) %>%
  filter(time <= time_outcome) %>%
  mutate(time_expo=time-time_outcome) %>% 
  ungroup()

```


### Description of the sample

```{r}

expo <- paquid_expo %>% group_by(ID) %>% arrange(time) %>%
  filter(row_number()==1)
summary(expo)

summary(paquid_outcome)

```

## Models considerd

First, We consider the following linear mixed model, where $i$
is the subject and $j$ the repeated measurement:

$$ normMMSE_{ij}= \beta_0+\beta_1time_{ij}+\beta_2sex_{i} + b_{0i}+b_{1i}time_{ij}+\varepsilon_{ij} $$
Where $b_{i}\sim \mathcal(N)(0,$B) and $\varepsilon_{ij}\sim N(0,\sigma^2)$

Next, we model the occurrence of physical dependency (HIER2 ≥ 2) using a weighted cumulative exposure (WCE) logistic regression. The model is written as:

$$ logit(P(HIER_i=1))=\alpha_0+\alpha_1sex_i+\sum_{k=0}^K\delta_k\sum_{t=-S}^0B_k(t)\times normMMSE_i^*(t)$$


where $B_k(t)$ is a basis of natural cubic splines with $K-1$ chosen internal knots.


## Estimate the exposure effect with 0 knots

To begin, we need to fit a linear mixed-effects exposure model using the $hlme$ function from the $lcmm$ package. For the package to work correctly later (e.g., in WCEland), it's important to model the time variable using flexible time functions such as natural cubic splines.
In our case, we chose to use natural splines with 4 internal knots. The number of internal knots was selected based on the Akaike Information Criterion (AIC) from different mixed linear models.

```{r, message=F}
library(lcmm)
library(splines)
library(mixWCE)
library(ggplot2)
m_expo<-hlme(normMMSE~ns(time_expo, knots = c(-16,-12,-8,-4),Boundary.knots = c(-20,-0))+male,
             random = ~ns(time_expo, knots = c(-16,-12,-8,-4),Boundary.knots = c(-20,-0)),
             subject = "ID",data = paquid_expo,returndata = T)

m_outcome<-WCEland(mexpo = m_expo,var.time = "time_expo",time.frame = c(-20,0,1),weightbasis = "NS"
              ,knots.vector = list(knots=NULL, boundary.knots=c(-20,-0)),data = paquid_outcome
              ,reg.type = "logistic",model = HIER2~male+WCIE,n_boot = 20)

```

## Estimate the exposure effect with more than 0 knots

Similarly, the number of internal knots used to estimate the exposure effect (i.e., the weighting function) in the outcome model is also selected based on the Akaike Information Criterion (AIC).
We compare several models with different knot configurations and retain the one that provides the best fit — that is, the model with the lowest AIC — for the binary health outcome of interest.

```{r, message=F}

m_outcome2<-WCEland(mexpo = m_expo,var.time = "time_expo",time.frame = c(-20,0,1),weightbasis = "NS"
              ,knots.vector = list(knots=c(-10), boundary.knots=c(-20,-0)),data = paquid_outcome
              ,reg.type = "logistic",model = HIER2~male+WCIE,n_boot = 20)


m_outcome3<-WCEland(mexpo = m_expo,var.time = "time_expo",time.frame = c(-20,0,1),weightbasis = "NS"
              ,knots.vector = list(knots=c(-13,-7), boundary.knots=c(-20,-0)),data = paquid_outcome
              ,reg.type = "logistic",model = HIER2~male+WCIE,n_boot = 20)

m_outcome$AIC
m_outcome2$AIC
m_outcome3$AIC

```
In model selection, a lower AIC suggests a better-fitting model. Therefore, we selected the model with 0 internal knots.

## Description of the WCE model

### summary of the model

The output above presents the full summary of the WCEland model fit. It includes:

- **Parameter Estimates (bootstrap-based):** Coefficients of the outcome logistic model estimated using a weighted bootstrap approach. These include the estimated coefficients, standard errors, Wald test statistics, and p-values.

- **Goodness-of-fit statistics:** Global fit metrics such as the maximum log-likelihood and the Akaike Information Criterion (AIC), which help evaluate and compare model adequacy.

- **Variance-covariance matrix:** Estimated variance-covariance matrix of the model parameters, useful for inference and understanding parameter correlations.

- **Weighted cumulative exposure (WCE) effect summary:** A table showing the time-specific estimated effects of past exposure on the outcome. For each lag time, the table includes the effect estimate, its standard error, the Wald statistic, and the associated p-value. This highlights how the influence of the exposure evolves over time.

- **Global WCE effect summary:** Mean cumulative effect across time lags, its standard error, the overall Wald statistic, and p-value — providing a concise summary of the exposure effect over the entire time window.

```{r}
summary(m_outcome)
```

### plot the expsoition trajectory association

The figure displays the estimated **weighted cumulative exposure (WCE) effect** trajectory over time. Each point on the x-axis represents a specific time lag preceding the outcome, while the y-axis shows the estimated effect of exposure at that time on the outcome.

- The **solid black curve** shows the estimated time-varying exposure effect.
- The **shaded area** represents the 95% confidence interval around the estimate, obtained through bootstrap replicates and by using a delta method.
- The **horizontal line at zero** helps visualize whether the estimated effect is statistically different from no effect.

```{r}
plot(m_outcome)
```
This plot displays the estimated contribution of past exposure to the **log-odds** of experiencing the outcome, assuming exposure levels at all other time points are held constant.
