---
title: "Estimate weight cumulative exposure effect for a survival outcome"
output: 
  rmarkdown::html_vignette:
    toc: true #table of content true
    toc_depth : 3
vignette: >
  %\VignetteIndexEntry{Estimate weight cumulative exposure effect for a survival outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette illustrates how to estimate weighted cumulative effect of a longitudinal exposition following the methodology described in Wagner et al. (2021 - https://doi.org/10.1186/s12874-021-01403-w). 

```{r, results='hide',message=FALSE,warning=FALSE, echo=FALSE}
# load("models_IRT.RData")
```
# Importation

The longitudinal IRT model is estimated via WCIE2F function of mixWCE R package. 
The following libraries are used in this vignette:

```{r,message=F}
library(lcmm)
library(ggplot2)
library(splines)
library(survival)
library(dplyr)
```



# Dataset (à revoir)

The illustration is on a simulated dataset that mimics the PREDIALA study described and analyzed in Proust-Lima et al. (2021). The dataset is called simdataIRT. It contains the following information:

- grp: group with 0=dialyzed and 1=preemptive
- sex: sex with 0=woman and 1=man
- age: age at entry in the cohort
- hads_2 ... hads_14 : 7 items of HADS measuring depression
- ID: identification number of the patient
- time: time of measurement (months since entry on the waiting list)
- time_entry: time on the waiting list at entry in the cohort (in months)

```{r}
str(simdataHADS)
```

## Description of the sample

```{r}
demo <- simdataHADS %>% group_by(ID) %>% arrange(time) %>%
  filter(row_number()==1)
summary(demo)
```


